Should geophysicists use gravity anomaly or gravity disturbance?

### 2018-05-10 manuscript/letter.tex
small changes suggested by Val

In a local or regional study, geophysicists commonly use a topocentric 
Cartesian coordinate system (TCS) with origin at a point $P$ on or 
close to the Earth's surface and axes $x$, $y$ and $z$ (Fig. 
\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
whereas the $z$-axis is opposite to the unit vector 
$\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) and points downward.
Consider, for example, a TCS (Fig. \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Fig. \ref{fig:GCS-GGS}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is a $3 \times 3$ orthogonal matrix whose 
first, second and third columns are defined, respectively, 
by the unit vectors $\hat{\mitbf{v}}$, $\hat{\mitbf{w}}$ 
and $-\hat{\mitbf{u}}$ (eq. \ref{eq:unit-vectors}), 
evaluated at the point $P$.

@@

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
Because the centrifugal part of the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
It is important to stress that both density distributions are unknown.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between the magnitudes of gravity vector
$g_{P} = \| \mitbf{g}_{P} \|$ and the normal gravity vector
$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$, at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
Another anomalous quantity is the \textit{gravity anomaly vector}.
It is defined, at point $P$, as the difference between the gravity
vector at a point $Q^{\prime}$ on the Geoid (a particular
equipotential surface of the gravity potential) and the normal
gravity vector at a point $Q$ on the ellipsoid surface, both at the
same geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
\begin{equation}
\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime}} - \mitbf{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}
Similarly to the gravity disturbance expression (eq.
\ref{eq:gravity-disturbance}), the gravity anomaly is given by
\begin{equation}
\Delta g_{P} = g_{Q^{\prime}} - \gamma_{Q} ,
\label{eq:gravity-anomaly}
\end{equation}
where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
the magnitude of the gravity vector $\mitbf{g}_{Q}$ on the Geoid,
at $Q^{\prime}$, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ 
on the reference ellipsoid, at the point $Q$ (Fig. \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have
implicitly considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that is allowed to be upward continued.

@@

The simple Bouguer anomaly is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
It is defined, over the continents, as follows 
\citep{blakely1996, hofmann-wellenhof-moritz2005}:
\begin{equation}
\Delta g_{P}^{B} 
= \Delta g_{P}^{F} - 2 \pi G \rho_{t} \, H_{P} \: ,
\label{eq:simple-bouguer-anomaly}
\end{equation}
where $\Delta g_{P}^{F}$ is the Free-air anomaly (eq. \ref{eq:free-air-anomaly})
and the last term on the right side is the simple Bouguer correction.
This term approximates the gravitational attraction that all topographic 
masses above the Geoid exerts at the point $P$ by the gravitational 
attraction of a homogeneous, infinitely extended slab of 
constant density $\rho_{t}$ and thickness equal to $H_{P}$, which is 
the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the point $P$.
By computing the simple Bouguer correction at a set of points
of a survey around $P$, we approximately remove, 
from the gravity measured at the points of the survey,
the gravitational effect of a homogeneous model representing 
the topographic masses above the Geoid.
It is evident from eqs \ref{eq:free-air-anomaly}, \ref{eq:gravity-disturbance-approx}
and \ref{eq:difference-disturbance-free-air} that the simple Bouguer
anomaly $\Delta g_{P}^{B}$ (eq. \ref{eq:simple-bouguer-anomaly}) represents
the gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
minus the gravitational effect produced by this homogeneous topographic model.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

There was a certain lack of comprehension regarding the
geophysical meaning of gravity anomalies until the
mid 90's.
As properly pointed out by \citet{chapin1996} at that time, 
``although the corrections which bring about a Bouguer 
gravity anomaly are well established, the reasons for doing
them are not well understood. One cause of this common 
misunderstanding is that the subject has been poorly presented in
many of the basic texts".
In his seminal book, \citet{blakely1996} brought some light
on the geophysical meaning of gravity anomalies from the 
perspective of applied geophysics. \citet{blakely1996} correctly
defined gravity sources as density contrasts between the actual
internal density distribution of the Earth and the internal density
distribution of the normal Earth.
However, he did not stress that, by removing the normal gravity 
evaluated on the ellipsoid from the gravity measured 
on the Earth's surface, the remaining disturbing field will reflect 
not only the effect produced by the gravity sources, but also a
small combination of gravitational and centrifugal effects.
This additional, non-harmonic and undesired effect is 
simply due to the calculation of the normal gravity at a point
other than that were the gravity is measured.

>>>>

In local or regional studies, geophysicists commonly use a topocentric 
Cartesian coordinate system (TCS) with origin at a point $P$ on or 
close to the Earth's surface and axes $x$, $y$ and $z$ (Fig. 
\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
whereas the $z$-axis is opposite to the unit vector 
$\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) and points downward.
Consider, for example, a TCS (Fig. \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Fig. \ref{fig:GCS-GGS}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is a $3 \times 3$ orthogonal matrix whose 
first, second and third columns are defined, respectively, 
by the unit vectors $\hat{\mitbf{v}}$, $\hat{\mitbf{w}}$ 
and $-\hat{\mitbf{u}}$ (eq. \ref{eq:unit-vectors}), 
evaluated at the point $P$.

@@

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
Because the centrifugal part of the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational effect caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
ergo, the gravity disturbance vector $\delta \mitbf{g}_{P}$ 
(eq. \ref{eq:gravity-disturbance-vector}) is harmonic.
It is important to stress that both density distributions are unknown.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between the magnitudes of gravity vector
$g_{P} = \| \mitbf{g}_{P} \|$ and the normal gravity vector
$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$, at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.

On the other hand, the \textit{gravity anomaly vector} at $P$ 
is the difference between the gravity
vector at a point $Q^{\prime}$ on the Geoid (a particular
equipotential surface of the gravity potential) and the normal
gravity vector at a point $Q$ on the ellipsoid surface, both at the
same geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
\begin{equation}
\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime}} - \mitbf{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}
Similarly to the gravity disturbance expression (eq.
\ref{eq:gravity-disturbance}), the gravity anomaly is given by
\begin{equation}
\Delta g_{P} = g_{Q^{\prime}} - \gamma_{Q} ,
\label{eq:gravity-anomaly}
\end{equation}
where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
the magnitude of the gravity vector $\mitbf{g}_{Q}$ on the Geoid,
at $Q^{\prime}$, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ 
on the reference ellipsoid, at the point $Q$ (Fig. \ref{fig:surfaces}).
As properly pointed out by \citet{barthelmes2013} the gravity anomaly,
by definition, depends on 
longitude and latitude only and is not a function of height. 
Because the gravity is on the Geoid and the normal gravity 
is on reference ellipsoid, the gravity anomaly $\Delta g_{P}$ 
(eq. \ref{eq:gravity-anomaly}) is not a purely gravitational effect 
due to the gravity sources, but rather a combination of gravitational 
and centrifugal effects. Consequently, its upward continuation, 
for example, could not be rigorously done.

However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have
implicitly considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function and can then being upward continued.

@@

The simple Bouguer anomaly is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
It is defined, over the continents, as follows 
\citep{blakely1996, hofmann-wellenhof-moritz2005}:
\begin{equation}
\Delta g_{P}^{B} 
= \Delta g_{P}^{F} - 2 \pi G \rho_{t} \, H_{P} \: ,
\label{eq:simple-bouguer-anomaly}
\end{equation}
where $\Delta g_{P}^{F}$ is the Free-air anomaly (eq. \ref{eq:free-air-anomaly})
and the last term on the right side is the simple Bouguer correction.
This term approximates the gravitational attraction that all topographic 
masses above the Geoid exerts at the point $P$ by the gravitational 
attraction of a homogeneous, infinitely extended slab of 
constant density $\rho_{t}$ and thickness equal to $H_{P}$, which is 
the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the point $P$.
By computing the simple Bouguer correction at a set of points
of a survey around $P$, we approximately remove, 
from the gravity measured at the points of the survey,
the gravitational effect of a homogeneous model representing 
the topographic masses above the Geoid.
It is evident from eqs 
\ref{eq:free-air-anomaly} to \ref{eq:difference-disturbance-free-air} 
that the simple Bouguer anomaly $\Delta g_{P}^{B}$ (eq.
\ref{eq:simple-bouguer-anomaly}) approximates the gravity disturbance
$\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
minus the gravitational effect produced by this homogeneous topographic
model.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

There was a certain lack of comprehension regarding the
geophysical meaning of gravity anomalies until the
mid 90's.
As properly pointed out by \citet{chapin1996} at that time, 
``although the corrections which bring about a Bouguer 
gravity anomaly are well established, the reasons for doing
them are not well understood. One cause of this common 
misunderstanding is that the subject has been poorly presented in
many of the basic texts".
In his seminal book, \citet{blakely1996} brought some light
on the geophysical meaning of gravity anomalies from the 
perspective of applied geophysics. \citet{blakely1996} correctly
defined gravity sources as density contrasts between the actual
internal density distribution of the Earth and the internal density
distribution of the normal Earth.
However, he did not stress that, by removing the normal gravity 
evaluated on the ellipsoid from the gravity measured 
on the Earth's surface, the remaining anomalous field will reflect 
not only the effect produced by the gravity sources, but also a
small combination of gravitational and centrifugal effects.
This additional, non-harmonic and undesired effect is 
simply due to the calculation of the normal gravity at a point
other than that were the gravity is measured.
